---
title: "Final_project"
output: html_document
date: "2024-03-07"
---

## Process error modeling

### Density independent model function

```{r}
load("~/Library/Mobile Documents/com~apple~CloudDocs/Documents/QSCI-454/harborseals.rda")

process.di.nll <- function(pars, nt){
  r <- pars[1]
  n_0 <- pars[2]
  nt.hat <- rep(NA, length(nt))
  n.loops <- length(nt)
  nt.hat[1] <- n_0 + n_0*r
  for (i in 2:n.loops){
    nt.hat[i]<- nt[i-1] +nt[i-1]*r
  }
  return(-sum(dpois(x=nt, lambda=nt.hat,log=TRUE)))
}

```

### Estimate r and N0

```{r}

nt.obs <- harborseals$abundance
start.pars <- c(0.05,2000)
process.di.fit <- optim(par = start.pars,
                        fn = process.di.nll,
                        nt = nt.obs,
                        method = "BFGS")

end_pars <- process.di.fit$par
di_nll <- process.di.fit$value

```

### Density dependent model function

```{r}

process.dd.nll <- function(pars, nt){
  r <- pars[1]
  n_0 <- pars[2]
  k <- pars[3]
  nt.hat <- rep(NA, length(nt))
  nt.hat[1] <- n_0 + n_0*r*(1 - (n_0/k))
  n.loops <- length(nt)
  for (i in 2:n.loops){
    nt.hat[i]<- nt[i-1] +nt[i-1]*r*(1-nt[i-1]/k)
  }
  return(-sum(dpois(x=nt, lambda=nt.hat,log=TRUE)))
}

```

### Estimate r, N0, K

```{r}

start.parsddp <- c(0.275, 3375, 8124)
process.dd.fit <- optim(par = start.parsddp,
                        fn = process.dd.nll,
                        nt = nt.obs,
                        method = "BFGS")

end_parsddp <- process.dd.fit$par
ddp_nll <- process.dd.fit$value

```

### Plot process error models (ask what Nt-1 is for the process error, think it's N0 but not sure)

```{r}
#functions to produce model values while also saving their respective nt-1's (I could've done this with the earlier function but didn't know if it would screw up optim)
hdi.p <- function(pars, nt){
  r <- pars[1]
  n_0 <- pars[2]
  nt.hat <- matrix(NA, nrow = length(nt), ncol=2)
  n.loops <- length(nt)
  nt.hat[1,1] <- n_0 + n_0*r
  nt.hat[1,2] <- n_0
  for (i in 2:n.loops){
    nt.hat[i,2] <- nt[i-1]
    nt.hat[i,1]<- nt[i-1] +nt[i-1]*r
  }
  return(nt.hat)
}
  
hdd.p <- function(pars, nt){
  r <- pars[1]
  n_0 <- pars[2]
  k <- pars[3]
  nt.hat <- matrix(NA, nrow = length(nt), ncol=2)
  nt.hat[1,1] <- n_0 + n_0*r*(1 - (n_0/k))
  nt.hat[1,2] <- n_0
  n.loops <- length(nt)
  for (i in 2:n.loops){
    nt.hat[i,2] <- nt[i-1]
    nt.hat[i]<- nt[i-1] +nt[i-1]*r*(1-nt[i-1]/k)
  }
  return(nt.hat)
}

#create list of nt-1 to pair with nt for graphing purposes
nt.1 <- rep(NA, length(nt.obs))
nt.observed <- data.frame(nt.obs, nt.1)
for (i in 2:length(nt.observed$nt.obs)){
  nt.observed[i, 2] <- nt.observed[i-1,1]
}

#create nt and nt-1 matrices for graphing for each model
hdi.p1 <- hdi.p(end_pars,nt.obs)
hdd.p1 <- hdd.p(end_parsdd, nt.obs)
 
# plot models and data  
plot(nt.observed$nt.1, nt.observed$nt.obs,
     type="p",
     xlab = "Nt-1",
     ylab = "Nt")
lines(hdi.p1[,2], hdi.p1[,1] , lwd = 2, col = "darkblue")
lines(hdd.p1[,2], hdd.p1[,1], lwd = 2, col = "darkred")
legend("topleft", lwd = 2, col = c("darkblue","darkred"), legend = c("Density Ind.","Density dep."))

```
## Observation error modeling

### Density independent model function

```{r}
obs.di.nll <- function(pars, nt.obs){
  r <- pars[1]
  n_1 <- pars[2]
  nt <- rep(NA, length(nt.obs))
  nt[1] <- n_1
  n.loops <- length(nt.obs)
  for (i in 2:n.loops){
    nt[i]<- nt[i-1] +nt[i-1]*r
  }
  return(-sum(dpois(x=nt.obs, lambda=nt,log=TRUE)))
}

```

### Estimate r and N1

```{r}
start.parsdio <- c(0.04, 4560)
obs.di.fit <- optim(par = start.parsdio,
                        fn = obs.di.nll,
                        nt = nt.obs,
                        method = "Nelder-Mead")

end_parsdio <- obs.di.fit$par
dio_nll <- obs.di.fit$value

```

### Density dependent model function

```{r}
obs.dd.nll <- function(pars, nt.obs){
  r <- pars[1]
  n_1 <- pars[2]
  K <- pars[3]
  nt <- rep(NA, length(nt.obs))
  nt[1] <- n_1
  n.loops <- length(nt.obs)
  for (i in 2:n.loops){
    nt[i]<- nt[i-1] +nt[i-1]*r*(1-nt[i-1]/K)
  }
  return(-sum(dpois(x=nt.obs, lambda=nt,log=TRUE)))
}

```

### Estimating r, N1, K

```{r}
start.parsddo <- c(0.18, 3848, 9377)
obs.dd.fit <- optim(par = start.parsddo,
                        fn = obs.dd.nll,
                        nt = nt.obs,
                        method = "Nelder-Mead")

end_parsddo <- obs.dd.fit$par
ddo_nll <- obs.dd.fit$value

```
### Plot observation error model

```{r}
t <- 1:length(nt.obs)

HDIO <- rep(NA, length(nt.obs))
rdio.mle <- end_parsdio[1]
HDIO[1] <- end_parsdio[2]
for (i in 2:length(nt.obs)){
  HDIO[i]<- HDIO[i-1] +HDIO[i-1]*rdio.mle
}
  
HDDO <- rep(NA, length(nt.obs))
rddo.mle <- end_parsddo[1]
HDDO[1] <- end_parsddo[2]
kddo.mle <- end_parsddo[3]
for (i in 2:length(nt.obs)){
  HDDO[i]<- HDDO[i-1] +HDDO[i-1]*rddo.mle*(1-HDDO[i-1]/kddo.mle)
}
  

# plot models and data  
plot(t,nt.obs ,
     type="p",
     xlab = "Time",
     ylab = "Nt")
lines(t, HDIO, lwd = 2, col = "darkblue")
lines(t, HDDO, lwd = 2, col = "darkred")
legend("topleft", lwd = 2, col = c("darkblue","darkred"), legend = c("Density Ind.","Density dep."))

```

## AIC table

```{r}
AIC_table <- data.frame(
  Model = c("Process DI", "Process DD", "Obs DI", "Obs DD"),
  Num_est_param = c(2, 3, 2, 3),
  NLL = c(di_nll, ddp_nll, dio_nll, ddo_nll)
)

AIC_table$AIC <- NA
for (i in 1:4){
  AIC_table$AIC[i] <- 2*(AIC_table$NLL[i] + AIC_table$Num_est_param[i])
}

AIC_table$deltAIC <- NA
for (i in 1:4){
  AIC_table$deltAIC[i] <- AIC_table$AIC[i] - min(AIC_table$AIC)
}

print(AIC_table)
```

## Confidence intervals for r, likelihood profile with K
```{r}

rlist <- seq(.025,.525,0.005)
klist <- seq(7624,8624,10)
nll <- matrix(data = NA, nrow=length(rlist), ncol=length(klist))
for (i in 1:length(rlist)){
  r <- rlist[i]
  for (j in 1:length(klist)){
    k <-klist[j]
    pars <- c(r, 3375.001, k)
    nll[i,j] <- process.dd.nll(pars, nt.obs)
  }
}

r.profile <- apply(X = nll, MARGIN = 1, FUN=min)
plot(rlist, r.profile, type = "l", lwd = 2)
min(r.profile)
r.profile
# min nll is element 16 in r.profile, so it's element 16 in rlist

rlist[51]

#value of r that maximizes likelihood is 3.5

confintnb <- min(r.profile)+1.92
r.profile

# index values for confidence interval looking at r list are 47 and 55
rlist[47]
rlist[55]

# conf int r list values are between 0.255 and 0.295
# r values less than 1 don't support complex pop dynamics in density dependent model

```




