---
title: "Lab_1"
output: html_document
date: "2024-01-03"
---

Learning R basics
```{r}
#scalars
a <- 3
b <- 2
a + b
#array
#c() command means to combine
arrie <- c(1,2,3,4)
# seq = create a sequence of numbers 
barry <- seq(from =1, to = 4, by=1)
#: makes list of integers
carrie <- 1:4
print(carrie)

#View makes matrix look like spreadsheet
View(carrie)

#indexing
carrie[3] 
carrie[3] <- 30

#loops
#assign numbers to objects, best practice in programming
#loop to create list of integers from 0-29
n.2.count <- 30
count <- rep(NA, n.2.count)
count[1] <- 0
for (i in 2:n.2.count){
  count[i] <- count[i-1] + 1
}
```

Practicum:
I will create the objects (N0, r, K, years) I need and assign values to them using arrows. I will create an array called "pops" by using the rep command, placing enough NA's in there that correspond to the # of years we want to run the simulation (contained in object "years"). For my loop, I will say for (i in 1:years), I will get the pop value for the next year by adding the previous year (Nt, pops[i-1]) to the previous year times r, and then times (1-Nt/K). I will assign the first value of the pops array (pops[1]) to be equal to N0 and start the loop at i = 2. I will graph the model output with year on the x axis and pop on the y axis. The code worked if the pop doesn't go over carrying capacity (100) and there are no NA values in the pops array. 

```{r}
#creating a loop to calculate wolf population 
#current wolf population size is 25, r equals 0.15, and K equals 100, 
#simulate the population size for the next 20 years

#create the variables
N0 <- 25
r <- .15
K <- 100
numyears <- 20
pops <- rep (NA, numyears)
#set first value for population
pops[1] <- N0

#run loop 
for (i in 2:numyears){
  n.t <- pops[i-1]
  n.t.plus1 <- n.t + r*n.t*(1-(n.t/K))
  pops[i] <- n.t.plus1
}

#see what pop values are
pops

#graph pops over years
years <- 1:20
plot(x= years, y = pops,
     xlab = "Years",
     ylab = "Wolf population")
```



